---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# img2ASCII

The goal of img2ASCII is to convert images (JPEG or PNG) into ASCII text representations. It provides both underlying functions for conversion and a simple Shiny webb application for interactive use.

## Installation

You can install the development version of img2ASCII from [GitHub](https://github.com/) with:
``` r
# install.packages("remotes")
remotes::install_github("ScottAustinYoung/img2ASCII.R")
```

## Example

The primary way to use this package is via the included Shiny application.

```{r example-usage, eval=FALSE}
library(img2ASCII)
launch_img2ascii_app()
```

In this app you can:
  - Upload a JPG or PNG image.
  - Adjust the maximum width of the generated ASCII array.
  - Adjust font size for the preview window.
  - Download the result as a text file or as an image.
  
## License
This package is licensed under the terms of the GNU General Public License v3.0.
